Unpaired measurement integration using Markov Link Intervals

### 2019-04-18 paper/paper.tex
Typos

Unfortunately, the $N_L<N_X$ regime is precisely the one which motivated our present work, inspired by investigation of cell-types in neuroscience.  This result shows that perfect estimation of $\beta$ may not be possible in our case.

@@

\hfill{}\begin{minipage}{.9\textwidth}
\caption{\textbf{Cell-type science: maximum likelihood estimation of $\beta$ suggests unlikely results about cell biology; Markov Link Intervals indicate plausible bounds.}  \label{fig:allen}
%
We performed unpaired measurement integration for two ways of measuring the `type' of mouse cells.  The first method uses gene expression to obtain a definitive measurement of the cell-type, $X$.  The second method yields a considerably less reliable estimate for the cell-type, $Y$.  For each cell we also know the value of $L$, which corresponds to another noisy indicator about the cell-type $X$.  In one dataset we can observe $(L,X)$ for many cells; in another dataset we can observe $(L,Y)$ many other cells.  Using this data we construct three estimates for $\beta$, the matrix of parameters which defines the relationship between $X$ and $Y$.  For each possible index $(x,y)$, $\hat \beta^{-}_{x,y}$ estimates a lower-bound estimate for $\beta_{x,y}$, $\hat \beta_{x,y}$ is a maximum likelihood estimator, and $\hat \beta^{+}$ estimates an upper bound.  Due to identifiability issues $\hat \beta_{x,y}$ is not guaranteed to be consistent.  However, Theorem \ref{thm:consistency} shows that the Markov Link Interval $[\hat \beta^{-}_{x,y},\hat \beta^{+}_{x,y}]$ is asymptotically guaranteed to include the truth.  Above we show all three of these estimators for a selection of values of $(x,y)$; we visualize these values in a grid, where rows correspond to values of $x$, columns correspond to values of $y$, and the color of the square in the $(x,y)$ cell indicates the value of the corresponding parameter.  The values of $\hat \beta$ are essentially nonsensical in many cases.  On the other hand, the Markov Link Interval suggests plausible bounds for the parameters and also indicates further experiments which would help refine these bounds.
}
\end{minipage}\hfill{}
\end{minipage}}\end{figure}

@@

For any cell it is thus impossible to measure a rich variety of cell properties \emph{and} reliably estimate the cell-type.  To understand the relationship between cell-type and these other cell properties, we therefore need some form of unpaired measurement integration.  This integration would allow us to determine the reliability of the noisy estimates $Y$, and thereby determine whether the relationship between cell-type and other cell properties can be reliably determined from the Patch-seq approach.  To facilitate this integration, we can construct experiments which make restrictions on which cells are are picked out, using a Cre/Lox system \cite{tasic2017shared}.  In each experiment, the probability that a particular cell will be picked out is entirely determined by the cell type $X$, and is thus statistically isolated from the noisy cell-type estimate $Y$.  By varying the experimental parameters we can change the distribution on which cells are picked out.  Letting $L$ denote these corresponding experimental parameters, we see that $L$ can act as a noisy independent indicator about the cell-type.  Markov Link Intervals should therefore give asymptotically consistent bounds on the reliability of the noisy estimate $Y$.

The Allen Institute recently conducted a series of experiments using both traditional RNA sequencing as well as the Patch-seq technique \cite{tasic2017shared}.  The same Cre/Lox systems were used for both techniques, thereby allowing us to observe the indicator $L$ for each sample in both datasets.  For each sample in the traditional RNA sequencing dataset (``Standard'') the cell-type was determined.  For each sample in the Patch-seq dataset we also have an estimated cell-type, although the Patch cell-type estimates are only given at a coarse level of the taxonomy of cells.  For example, in the Standard dataset a distinction is made between cell-type `Lamp5 Egln 1' and cell-type `Lamp5 Pdlim5,' whereas in the Patch-seq dataset we group all GABAergic neuronal cell-types characterized by high expression of `Lamp5' into a single coarse `cell class.'

We applied the Markov Link Intervals to the Allen Institute data try to determine the reliability of the noisy cell-type estimators $Y$ arising from Patch-seq approach.  We can visualize a maximum likelihood estimator as well as the Markov Link Intervals through heatmaps (see Figure \ref{fig:allen})).  We see that the maximum likelihood estimator sometimes gives essentially nonsensical results.  For example, the maximum likelihood estimator suggests that a particular group of excitatory glutamatergic neurons from visual cortex are \emph{always} assigned to a GABAergic inhibitory type by the $Y$ variable.  This is quite unlikely, since even the noisy measurements of $Y$ can reliably determine between excitatory and inhibitory neurons \cite{cadwell2016electrophysiological,cadwell2017multimodal}.

The Markov Link Intervals yield plausible bounds on the true model parameters.  For example, the lower bounds of the Markov Link Interval suggests that a sizeable fraction of `Lamp5 Pdlim5' cells are indeed classified as `Lamp5' cells by the Patch method (note that the Patch method does not give fine-grained cell-type predictions, such as `Lamp5 Pdlim5', but instead only gives coarse class-level cell-type predictions, such as `Lamp5').  The upper bounds furthermore show that these cells are seldom classified as anything \emph{but} `Lamp5' cells or possibly the `unknown' type which the Patch method assigns to cells which it cannot identify.  For cells of type `Sst Myh8-1',  Markov Link Intervals give us considerably less certainty regarding the related type; the intervals indicate that the Patch method does \emph{not} classify these cells as `CR,' `L6b,' `Meis2,' `Sncg,' or `Vip' -- but it does not indicate  how they \emph{are} being classified.  At the extreme end, we have cells of type `Lamp5 Elgn3-1,' for which the Markov Link Intervals are completely vacuous.   These `Lamp5 Elgn3-1' cells might get perfectly classified or they may get poorly classified; the available data cannot tell us either way.

Parameters with large intervals suggest new experiments to perform in order to more closely determine the model parameters.  For example, the significant ambiguity for cells with standard type `Lamp5 Egln 1' suggests we need a series of experiments with more distinct Cre/Lox experimental parameters that pick out cells of this kind.  If we could find a Cre/Lox procedure that obtained many `Lamp5 Egln-1' cells but no cells which measure as Patch type `Pvalb,' this would show that the Standard `Lamp5 Egln 1' is not associated with Patch type `Pvalb.'  On the other hand, if we could find Cre/Lox parameters that select many `Lamp5 Egln 1' cells but \emph{only} cells with Patch type `Pvalb,' this would show the opposite.  If these additional experiments were performed, new Markov Link Intervals could be computed on the new data to determine what aspects of the link are still ambiguous.

>>>>

Unfortunately, the $N_L<N_X$ regime is precisely the one which motivated our present work, inspired by investigation of cell-types in neuroscience.  This result shows that consistent estimation of $\beta$ may not be possible in our case.

@@

\hfill{}\begin{minipage}{.9\textwidth}
\caption{\textbf{Cell-type science.  Maximum likelihood estimation of $\beta$ suggests unlikely results about cell biology; Markov Link Intervals indicate plausible bounds.}  \label{fig:allen}
%
We performed unpaired measurement integration for two ways of measuring the type of cells from mouse visual cortex.  The first method uses gene expression to obtain a definitive measurement of the cell-type, $X$.  The second method yields a considerably less reliable estimate for the cell-type, $Y$.  For each cell we also know the value of $L$, which corresponds to another noisy indicator about the cell-type $X$.  In one dataset we can observe $(L,X)$ for many cells; in another dataset we can observe $(L,Y)$ many other cells.  Using this data we construct three estimates for $\beta$, the matrix of parameters which defines the relationship between $X$ and $Y$.  For each possible index $(x,y)$, $\hat \beta^{-}_{x,y}$ estimates a lower-bound estimate for $\beta_{x,y}$, $\hat \beta_{x,y}$ is a maximum likelihood estimator, and $\hat \beta^{+}$ estimates an upper bound.  Due to identifiability issues $\hat \beta_{x,y}$ is not guaranteed to be consistent.  However, Theorem \ref{thm:consistency} shows that the Markov Link Interval $[\hat \beta^{-}_{x,y},\hat \beta^{+}_{x,y}]$ is asymptotically guaranteed to include the truth.  Above we show all three of these estimators for a selection of values of $(x,y)$; we visualize these values in a grid, where rows correspond to values of $x$, columns correspond to values of $y$, and the color of the square in the $(x,y)$ entry indicates the value of the corresponding parameter.  The values of $\hat \beta$ are nonsensical in many cases.  On the other hand, the Markov Link Interval suggests plausible bounds for the parameters and also indicates further experiments which would help refine these bounds.
}
\end{minipage}\hfill{}
\end{minipage}}\end{figure}

@@

For any cell it is thus impossible to measure a rich variety of cell properties \emph{and} reliably estimate the cell-type.  To understand the relationship between cell-type and these other cell properties, we therefore need some form of unpaired measurement integration.  This integration would allow us to determine the reliability of the noisy estimates $Y$, and thereby determine whether the relationship between cell-type and other cell properties can be reliably determined from the Patch-seq approach.  To facilitate this integration, one can conduct experiments which make restrictions on which cells are are picked out, using a Cre/Lox system \cite{tasic2017shared}.  In each experiment, the probability that a particular cell will be picked out is entirely determined by the cell type $X$, and is thus statistically isolated from the noisy cell-type estimate $Y$.  By varying the experimental parameters we can change the distribution on which cells are picked out.  Letting $L$ denote these corresponding experimental parameters, we see that $L$ can act as a noisy independent indicator about the cell-type.  Markov Link Intervals should therefore give asymptotically consistent bounds on the reliability of the noisy estimate $Y$.

The Allen Institute recently conducted a series of experiments using both traditional RNA sequencing as well as the Patch-seq technique \cite{tasic2017shared}.  The same Cre/Lox systems were used for both techniques.  For each sample in the traditional RNA sequencing dataset we observe both the cell-type ($X$) and the Cre/Lox indicator ($L$).  For each sample in the Patch-seq dataset we observe a both noisy estimate of the cell-type ($Y$) and the Cre/Lox indicator ($L$).  
% Note that the Patch cell-type estimates are only given at a coarse level of the taxonomy of cells.  For example, in the traditional RNA sequencing dataset a distinction is made between cell-type `Lamp5 Egln 1' and cell-type `Lamp5 Pdlim5,' whereas in the Patch-seq dataset we group all GABAergic neuronal cell-types characterized by high expression of `Lamp5' into a single coarse `cell class.'

We applied the Markov Link Intervals to the Allen Institute data try to determine the reliability of the noisy cell-type estimators $Y$ arising from Patch-seq approach.  We can visualize a maximum likelihood estimator as well as the Markov Link Intervals through heatmaps (see Figure \ref{fig:allen}).  We see that the maximum likelihood estimator sometimes gives nonsensical results.  For example, the maximum likelihood estimator suggests that a particular group of excitatory glutamatergic neurons from visual cortex are \emph{always} assigned to a GABAergic inhibitory type by the $Y$ variable.  This is quite unlikely, since even the noisy measurements of $Y$ can reliably determine between excitatory and inhibitory neurons \cite{cadwell2016electrophysiological,cadwell2017multimodal}.

The Markov Link Intervals yield plausible bounds on the true model parameters.  For example, the lower bounds of the Markov Link Interval suggests that a sizeable fraction of `Lamp5 Pdlim5' cells are indeed classified as `Lamp5' cells by the Patch method (note that the Patch method does not give fine-grained cell-type predictions, such as `Lamp5 Pdlim5', but instead only gives coarse class-level cell-type predictions, such as `Lamp5').  The upper bounds show that these cells are not commonly categorized into any group besides the `Lamp5' type and the the `unknown' type which the Patch method assigns to cells which it cannot identify.  Consider another example: for cells of type `Sst Myh8-1',  Markov Link Intervals give us considerably less certainty regarding the related type.  The intervals indicate that the Patch method does not typically classify these cells as `CR,' `L6b,' `Meis2,' `Sncg,' or `Vip' -- but it does not indicate how they are being classified.  At the extreme end, we have cells of type `Lamp5 Elgn3-1,' for which the Markov Link Intervals are completely vacuous.   It is possible that the Patch method perfectly classifies these `Lamp5 Elgn3-1' cells; it is possible that it always misclassifies them; the available data cannot tell us either way.

Parameters with large intervals suggest new experiments to perform in order to more closely determine the model parameters.  For example, the significant parameter ambiguity for cells with type `Lamp5 Egln 1' suggests we need a series of experiments with more distinct Cre/Lox experimental parameters that pick out cells of this kind.  If we could find a Cre/Lox procedure that obtained many `Lamp5 Egln-1' cells but no cells which Patch categorizes as `Pvalb,' this would show that `Lamp5 Egln 1' cells are not associated with Patch type `Pvalb.'  On the other hand, if we could find Cre/Lox parameters that select many `Lamp5 Egln 1' cells and also selects cells which are largely categorized as `Pvalb' by Patch, this would show the opposite.  If these additional experiments were performed, new Markov Link Intervals could be computed on the new data to determine what aspects of the link are still ambiguous.
